Pre-/Post-Trikafta in CFRD

### 2022-02-08 Christine Chan/Pre-post Trikafta/pre_post_trikafta_analysis.Rmd
Update pre_post_trikafta_analysis.Rmd

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(lmerTest)
library(nlme)
library(broom.mixed)
library(segmented)
library(performance)
library(lubridate)
library(plotly)
library(knitr)
knitr::opts_chunk$set(echo = FALSE,fig.height = 8,fig.width = 10)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "S:/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
```

```{r include=FALSE}
# Import cleaned data from Casey
df = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/analysis_data.csv",na.strings = "NA")
df$X = NULL
# Fill in MRN and CFF ID for everyone
df = df %>% 
  group_by(CFF.ID) %>% fill(MRN,.direction = "downup") %>% 
  group_by(MRN) %>% fill(CFF.ID,.direction = "downup")
# Assume that blanks for pancreatic and G tube indicate "no"
df$Hispanic.Latinx[is.na(df$Hispanic.Latinx)] = FALSE
df$Pancreatic.Status[is.na(df$Pancreatic.Status)] = "Sufficient"
df$G.tube.in.past.12.months[is.na(df$G.tube.in.past.12.months)] = "No"
# Convert dates
date_vars = c("DOB","Start","Date")
df[,date_vars] = lapply(df[,date_vars], function(d){
  parse_date_time(d,orders = c("mdy","ymd","dmy"),quiet = T)
})
# Time from start and filtering
df = df %>% mutate(Time = as.numeric(difftime(Date,Start,units = "days"))) 
# Add UM (they did not provide dates, so need Time column)
um_a1cs = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_a1cs.csv",na.strings = "")
um_bmi = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_bmi.csv",na.strings = "")
um_demo = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_demo.csv",na.strings = "")
um_ogtt = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_ogtt.csv",na.strings = "")
um_pfts = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_pfts.csv",na.strings = "")
um = full_join(um_a1cs,um_bmi)
um = full_join(um,um_ogtt)
um = full_join(um,um_pfts)
um = left_join(um,um_demo)
um = um %>% filter(!is.na(CFRD.Status))
df = full_join(df,um)
# Filter and fill constants
df = df %>% group_by(CFF.ID) %>% fill(DOB,Start:CFRD.Diagnosis.Date,.direction = "downup") %>%
  ungroup() %>% filter(!is.na(Time),abs(Time) <= 365,!is.na(CFF.ID),
                       (!is.na(CFRD.Diagnosis.Date) | !is.na(CFRD.Status) | site == "UW")) %>%
  select(-MRN) %>% select(CFF.ID,site,DOB,Start,everything()) %>% rename(Site = "site")

```

```{r}
data = df
outcome = "HbA1c"
# Piecewise model function
piecewise = function(outcome,data = df){
  # Add spline column
  data$Change = ifelse(data$Time < 0,0,1)
  data$Change = data$Change * data$Time
  # Model
  f = as.formula(paste0(outcome,"~Time + Change + (1|CFF.ID)"))
  mod = lmer(f,data = data)
  res = tidy(mod,"fixed")
  res$effect = NULL
  #res$term[2] = "Slope Prior"
  # Segmented fit
  fit1 = lme(as.formula(paste0(outcome,"~Time + Change")),random = ~1|CFF.ID,data = data,na.action = na.omit)
  seg = suppressWarnings(segmented(fit, seg.Z = ~Time, npsi=1))
  # Model check if necessary
  check = check_model(mod)
  # Coordinates for model plotting
  x1_start = -365
  y1_start = res$estimate[1] + res$estimate[2]*x1_start
  x1_end = 0
  y1_end = res$estimate[1] 
  x2_start = 0
  y2_start = res$estimate[1] 
  x2_end = 365
  y2_end = res$estimate[1] + (res$estimate[2]+res$estimate[3])*x2_end
  # Plot
  p = ggplot(data,aes_string(x = "Time",y = outcome,group = "CFF.ID",color = "Site")) +
    geom_point(alpha = 0.5) + geom_line(alpha = 0.2) +
    annotate("segment",x=x1_start,y=y1_start,xend=x1_end,yend=y1_end,color = "blue",size = 1) +
    annotate("segment",x=x2_start,y=y2_start,xend=x2_end,yend=y2_end,color = "blue",size = 1) +
    xlab("Days from Trikafta Start") + theme_bw()
  # Interactive plot and model results
  return(list(plot = ggplotly(p),table = res,check = check,seg = seg))
}
```

@@

```{r results='asis'}
bmi = piecewise(outcome = "BMI",df)
ggplotly(bmi$plot)
kable(bmi$table,digits = 3)
bmi$check
```

`r if(class(bmi$seg)=="lme"){"Automated R package was not able to find a better breakpoint."}`

>>>>

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
library(lmerTest)
library(nlme)
library(broom.mixed)
library(segmented)
library(performance)
library(AGD)
library(rspiro)
library(lubridate)
library(plotly)
library(knitr)
knitr::opts_chunk$set(echo = FALSE,fig.height = 8,fig.width = 10)
if(Sys.info()["sysname"] == "Windows"){
  home_dir = "S:/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
```

```{r include=FALSE}
# Import cleaned data from Casey
df = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/analysis_data.csv",na.strings = "NA")
df$X = NULL
# Fill in MRN and CFF ID for everyone
df = df %>% 
  group_by(CFF.ID) %>% fill(MRN,.direction = "downup") %>% 
  group_by(MRN) %>% fill(CFF.ID,.direction = "downup")
# Assume that blanks for pancreatic and G tube indicate "no"
df$Hispanic.Latinx[is.na(df$Hispanic.Latinx)] = FALSE
df$Pancreatic.Status[is.na(df$Pancreatic.Status)] = "Sufficient"
df$G.tube.in.past.12.months[is.na(df$G.tube.in.past.12.months)] = "No"
# Convert dates
date_vars = c("DOB","Start","Date")
df[,date_vars] = lapply(df[,date_vars], function(d){
  parse_date_time(d,orders = c("mdy","ymd","dmy"),quiet = T)
})
# Time from start and filtering
df = df %>% mutate(Time = as.numeric(difftime(Date,Start,units = "days"))) 
# Add UM (they did not provide dates, so need Time column)
um_a1cs = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_a1cs.csv",na.strings = "")
um_bmi = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_bmi.csv",na.strings = "")
um_demo = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_demo.csv",na.strings = "")
um_ogtt = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_ogtt.csv",na.strings = "")
um_pfts = read.csv("./Christine Chan/Prepost Triakfta/Data_Cleaned/um_pfts.csv",na.strings = "")
um = full_join(um_a1cs,um_bmi)
um = full_join(um,um_ogtt)
um = full_join(um,um_pfts)
um = left_join(um,um_demo)
um = um %>% filter(!is.na(CFRD.Status))
df = full_join(df,um)
# Filter and fill constants
df = df %>% group_by(CFF.ID) %>% fill(DOB,Start:CFRD.Diagnosis.Date,.direction = "downup") %>%
  ungroup() %>% filter(!is.na(Time),abs(Time) <= 365,!is.na(CFF.ID),
                       (!is.na(CFRD.Diagnosis.Date) | !is.na(CFRD.Status) | site == "UW")) %>%
  select(-MRN) %>% select(CFF.ID,site,DOB,Start,everything()) %>% rename(Site = "site")
df$Sex = as.factor(df$Sex)
levels(df$Sex) = c("F","M")
# BMI category
df$Age = pmax(as.numeric(difftime(df$Date,df$DOB,units = "days")),df$Age,na.rm = T)/365.25
df$BMIz = y2z(df$BMI,x = df$Age,sex = df$Sex,ref=nl4.bmi)
df$BMI_group = apply(df,1,function(r){
  age = as.numeric(r["Age"])
  if(!is.na(age) & age > 19){
    bmi = as.numeric(r["BMI"])
    group = cut(bmi,c(-Inf,18.5,25,30,35,40,Inf),
        labels = c("Under Weight","Normal Weight","Overweight","Obese (Class I)",
                   "Obese (Class II)","Obese (Class III+)"),right = F)
  } else if (!is.na(age) & age <= 19){
    z = as.numeric(r["BMIz"])
    perc = pnorm(z)
    group = cut(perc,c(-Inf,5,85,95,Inf),
        labels = c("Under Weight","Normal Weight","Overweight","Obese (Class I)"),right = F)
  } else {group = NA}
  return(as.character(group))
})
df$BMI_group = as.factor(df$BMI_group)
df$BMI_group = ordered(df$BMI_group,c("Under Weight","Normal Weight","Overweight",
                                      "Obese (Class I)","Obese (Class II)","Obese (Class III+)"))
# PFTs
old <- c("Caucasian","Black","Other")
new <- c(1,2,5)
df$ppFEV = pctpred_GLI(df$Age,df$Height/100,gender = as.numeric(df$Sex),
            ethnicity = new[match(df$Race, old)],
            FEV1 = df$FEV1)
df$ppFVC = pctpred_GLI(df$Age,df$Height/100,gender = as.numeric(df$Sex),
            ethnicity = new[match(df$Race, old)],
            FEV1 = df$FVC)
```

```{r}
data = df
outcome = "HbA1c"
# Piecewise model function
piecewise = function(outcome,data = df,ordinal = F){
  # Add spline column
  data$Change = ifelse(data$Time < 0,0,1)
  data$Change = data$Change * data$Time
  # Model
  f = as.formula(paste0(outcome,"~Time + Change + (1|CFF.ID)"))
  mod = lmer(f,data = data)
  res = tidy(mod,"fixed")
  res$effect = NULL
  # Fit ordinal GEE (for example BMI group)
  if(ordinal){
    f = as.formula(paste0(outcome,"~Time + Change"))
    mod = ordgee(f,data = data,id = data$CFF.ID)
    res = summary(mod)
  }
  # Segmented fit
  fit = lme(as.formula(paste0(outcome,"~Time + Change")),random = ~1|CFF.ID,
            data = data,na.action = na.omit)
  seg = suppressWarnings(segmented(fit, seg.Z = ~Time, npsi=1))
  # Model check if necessary
  check = check_model(mod)
  # Coordinates for model plotting
  x1_start = -365
  y1_start = res$estimate[1] + res$estimate[2]*x1_start
  x1_end = 0
  y1_end = res$estimate[1] 
  x2_start = 0
  y2_start = res$estimate[1] 
  x2_end = 365
  y2_end = res$estimate[1] + (res$estimate[2]+res$estimate[3])*x2_end
  # Plot
  p = ggplot(data,aes_string(x = "Time",y = outcome,group = "CFF.ID",color = "Site")) +
    geom_point(alpha = 0.5) + geom_line(alpha = 0.2) +
    annotate("segment",x=x1_start,y=y1_start,xend=x1_end,yend=y1_end,color = "blue",size = 1) +
    annotate("segment",x=x2_start,y=y2_start,xend=x2_end,yend=y2_end,color = "blue",size = 1) +
    xlab("Days from Trikafta Start") + theme_bw()
  # Interactive plot and model results
  return(list(plot = ggplotly(p),table = res,check = check,seg = seg))
}
```

@@

Models did not converge with BMI converted to under weight, normal weight, etc. This is most likely due to lack of within-patient variability in terms of BMI group (i.e. people's BMIs aren't changing enough that they move from one group to another).

# Insulin

Most participants only have one observation of insulin dose, so these results would be the same as Maxie's previous reports.

# ppFEV

```{r results='asis'}
ppFEV = piecewise(outcome = "ppFEV",df)
ggplotly(ppFEV$plot)
kable(ppFEV$table,digits = 3)
ppFEV$check
```

# ppFVC

```{r results='asis'}
ppFVC = piecewise(outcome = "ppFVC",df)
ggplotly(ppFVC$plot)
kable(ppFVC$table,digits = 3)
ppFVC$check
```
